---
title: "Research Question 2"
author: "Pingxuan Ren"
date: "2024-12-07"
output: pdf_document
---

```{r}
# Load the libraries
library(dplyr)
library(ggplot2)

# Import train Dataset
train = read.csv("train.csv")

# Fit a Generalized Linear Model
model_one <- glm(Diagnosis ~ BMI + Smoking + AlcoholConsumption + PhysicalActivity + DietQuality + SleepQuality, 
                 data = train, 
                 family = binomial(link = "logit"))

```

```{r}
# GLM Model analysis
summary(model_one)

# Extract variables by significance level (0.05)
significant_vars <- summary(model_one)$coefficients %>%
  as.data.frame() %>%
  filter(`Pr(>|z|)` < 0.05)

# Print significant variables
print(significant_vars)
```
After using GLM with respect to the six factors including BMI, Smoking, AlcoholConsumption, PhysicalActicity, DietQuality, and SleepQuality, the results of the analysis are shown below:
First of all, the null variance 1954.4 is slightly greater than the residual deviance of 1944.5, indicating that the health factors could explain a small proportion of variability in this training dataset. \
Secondly, according to our GLM model the only statistically significant predictor is SleepQuality with a P-value of 0.0121. This means that among the six health-related factors, sleep quality is beneficial for the treatment of Alzheimer. \
The odds ratio is:

\[
\text{OR} = e^{\beta_{\text{SleepQuality}}} = 0.925
\]

and this means that for every one-unit increase in the sleep quality among the subjects, the odds of being diagnosed with Alzheimer is multiplied by 0.925, or decrease by:

\[
\text{\%Change} = 1 - \text{Odds Ratio} = 1 - 0.925 = 7.5\%
\].
